---
title: "Diversity Analysis MS before vs after treatment, stool"
author: "Alba Troci"
date: "`r Sys.Date()`"  
output:
   BiocStyle::html_document:
      toc: true
      df_print: paged
      self_contained: true
      code_download: true 
      highlight: tango
editor_options:
  chunk_output_type: console
params:
  FIGPATH: "figures/"
  d.out: "./"
---
# Preparations
## Set global options

```{r style, results="asis", cache=FALSE, message = F, echo=FALSE}
# Set knit global options
library("knitr")
options(digits = 2, width = 80)
golden_ratio <- (1 + sqrt(5)) / 2
opts_chunk$set(echo = TRUE,
               tidy = FALSE,
               include = TRUE,
               fig.path = params$FIGPATH,
               dev=c("png",'pdf'),
               fig.height = 5,
               fig.width = 4 * golden_ratio,
               comment = '  ',
               dpi = 300,
               cache = FALSE)


# Pretty outputs
library("rmarkdown")
library("BiocStyle")
library("ggthemes") # add ggplot2 themes
library("ggplot2")
library("ggpubr")

# Set seed for reproducibility
set.seed(100)

# Set plotting theme
theme_set(theme_few(base_size = 14))

# Set output directory
d.out <- params$d.out
rm(params)
```


## Load libraries for the session
```{r libraries, message=FALSE}
library("magrittr")
library("ggplot2")
library("tidyverse")
library("readxl")
library("dada2")
library("phyloseq")
library("gridExtra")
library("vegan")
library("plyr")
library("scales")
library("reshape2")
library("DESeq2")
library("dplyr") #for data manipulation
library("msa") #multiple sequence alignment
library("ape") #to create phylogenetic tree
library("randomForest") 
library("caret") 
library("broom")
library("mlbench")
library("plotROC")
library("car")
library("MASS")
library("lme4")
library("lmerTest")
library("fitdistrplus")
library("psycho")
library("rstatix")
library("coin")
library("compareGroups")
library("rio")
```

## Import data: no agglomeration
```{r import data}
# Phyloseq object: Filtered ASV based from ST Samples
d.ps_ST1 <- "~/IKMB/projects/MS/github.codes.ms/ms.before.after/data"
ps_ST1 <- "ps.ms.stool.rds" %>%
  paste0(d.ps_ST1,"/", .) %>% 
  readRDS() 
ps_ST1

#Subset only treated samples
ps_ST <- subset_samples(ps_ST1, Ocrevus %in% "yes")
ps_ST

t <- data.frame(sample_data(ps_ST))

```


## Sample overview and filter

```{r}
#total nr of read counts of each sample
summary(sample_sums(ps_ST))
sample_sum_df <- data.frame(sum = sample_sums(ps_ST))

```

```{r}
#Data table
data2 <- data.frame(Sample = sample_data(ps_ST)$NewID, 
                   Participant = sample_data(ps_ST)$Samples,
                   TimePoint = sample_data(ps_ST)$TimePoint)
data2
tp1 <- subset(data2, TimePoint == "TP1")
```

```{r}
ggplot(data2, aes(x = TimePoint, y = Participant)) +
  geom_point()
```

## Edit
Now I will do diversity analysis based on Stool Samples,filtered, non agglomerated data.
Input: ps_ST

```{r, message=FALSE}
#Add a column named Group (Before=TP0,TP1 and After=TP2,TP3)
sample_data(ps_ST)$Group[sample_data(ps_ST)$TimePoint %in%
                           c("TP0", "TP1" )] <- "Before"
sample_data(ps_ST)$Group[sample_data(ps_ST)$TimePoint %in%
                           c("TP2", "TP3", "TP4" )] <- "After"
```

```{r}
#Ocrevus: yes (double check step)
ps_ST_oc <- subset_samples(ps_ST, Ocrevus == "yes")
ps_ST.TP2 <- subset_samples(ps_ST_oc, TimePoint %in% c("TP1", "TP2"))
dim(sample_data(ps_ST.TP2))
ps_ST.TP3 <- subset_samples(ps_ST_oc, TimePoint %in% c("TP1", "TP3"))
dim(sample_data(ps_ST.TP3))
ps_ST.TP4 <- subset_samples(ps_ST_oc, TimePoint %in% c("TP1", "TP4"))
dim(sample_data(ps_ST.TP4))

```

## Rarefaction: no agglomeration
We need to rarefy before alpha diversity:

```{r, message=FALSE}
set.seed(300)
summary(sample_sums(ps_ST.TP2))
summary(sample_sums(ps_ST.TP3))
summary(sample_sums(ps_ST.TP4))
```

```{r}
psST2.rar <- rarefy_even_depth(ps_ST.TP2, sample.size = min(sample_sums(ps_ST.TP2)))
psST2.rar

psST3.rar <- rarefy_even_depth(ps_ST.TP3, sample.size = min(sample_sums(ps_ST.TP3)))
psST3.rar

psST4.rar <- rarefy_even_depth(ps_ST.TP4, sample.size = min(sample_sums(ps_ST.TP4)))
psST4.rar
```

# Alpha diversity

Here I will use rarefied samples as an input, split them into Groups Treated and 
non Treated, then compute alpha diversity seperately.

## ST.TP2
```{r, message=FALSE}

# Richness (nr of species in on environment)
richness <- estimate_richness(psST2.rar, split = TRUE, measures = "shannon")
richness2 <- estimate_richness(psST2.rar, split = TRUE, measures = "Chao1")
richness3 <- estimate_richness(psST2.rar, split = TRUE, measures = "InvSimpson")

#Diversity values in one table only
rich <- estimate_richness(psST2.rar, measures = c("Shannon", "Chao1", "InvSimpson"))
head(rich)

```

## ST.TP3
```{r, message=FALSE}

# Richness (nr of species in on environment)
richness31 <- estimate_richness(psST3.rar, split = TRUE, measures = "shannon")
richness32 <- estimate_richness(psST3.rar, split = TRUE, measures = "Chao1")
richness33 <- estimate_richness(psST3.rar, split = TRUE, measures = "InvSimpson")

#Diversity values in one table only
rich3 <- estimate_richness(psST3.rar, measures = c("Shannon", "Chao1", "InvSimpson"))
head(rich3)

```

## ST.TP4
```{r, message=FALSE}

# Richness (nr of species in on environment)
richness41 <- estimate_richness(psST4.rar, split = TRUE, measures = "shannon")
richness42 <- estimate_richness(psST4.rar, split = TRUE, measures = "Chao1")
richness43 <- estimate_richness(psST4.rar, split = TRUE, measures = "InvSimpson")

#Diversity values in one table only
rich4 <- estimate_richness(psST4.rar, measures = c("Shannon", "Chao1", "InvSimpson"))
head(rich4)

```

# First approach: Fit a model
## Preparations
I will create two data.frames with Age,Sex,Group,TimePoint, Chao1 alpha diversity for Treated and non Treated and then merge together in one for further analysis.

### MS.TP1 vs MS.TP2
```{r}
#Prepare a data.frame for Treated group.

df1 = cbind(sample_data(psST2.rar)[, "Samples", drop=FALSE], richness2)
df1$Shannon <- richness$Shannon
df1$InvSimpson <- richness3$InvSimpson
df1$Age <- sample_data(psST2.rar)$Age
df1$Sex <- sample_data(psST2.rar)$Sex
df1$Group <- sample_data(psST2.rar)$Group
df1$TimePoint <- sample_data(psST2.rar)$TimePoint
df1$ParticipantID <- sample_names(psST2.rar)
df1$logChao1 <- log(df1$Chao1)
df1$Treat <- sample_data(psST2.rar)$Ocrevus
df1 <- df1[c(1,3,4,5,6,7,2,8,9,10,11,12)]
df1

#samples are in each time point
table(df1$TimePoint)
```

```{r}
#select only duplicated elements
xu1 <- df1$Samples[duplicated(df1$Samples)]
xu1

df11 <- subset(df1, Samples %in% xu1)
df11
```


#### Chao1

```{r}
#Check for normal distribution
shapiro.test(df11$Chao1)
hist(df11$Chao1,probability=T, main="MS.TP1 vs MS.TP2",xlab="Chao1") 
lines(density(df11$Chao1),col=2)

```


```{r}
test1 <- t.test(Chao1 ~ Group, data = df11, paired = TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)

test1

```

```{r}
##effect size
effect1 <- df11 %>% cohens_d(Chao1 ~ Group, var.equal = TRUE, hedges.correction = TRUE)
effect1

```

```{r}
#create tabl values

t1 <- data.frame(Group = "After(TP2) vs Before(TP1)", 
                 DataName = test1$data.name,
                 Method = test1$method,
                 After = effect1$n2, Before = effect1$n1, 
                 Statistic = test1$statistic, 
                 #df = test1$parameter, 
                 EffectSize = effect1$effsize,
                 Pvalue = test1$p.value)
```


#### Shannon

```{r}
shapiro.test(df11$Shannon)
hist(df11$Shannon,probability=T, main="MS.TP1 vs MS.TP2",xlab="Shannon") 
lines(density(df11$Shannon),col=2)

```


```{r}
test12 <- wilcox.test(Shannon ~ Group, data = df11, paired = TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)

test12

```

```{r}
##effect size
effect12 <- wilcox_effsize(df11, Shannon ~ Group, paired = TRUE)
effect12

```

```{r}
#create tabl values

t12 <- data.frame(Group = "After(TP2) vs Before(TP1)", 
                 DataName = test12$data.name,
                 Method = test12$method,
                 After = effect12$n2, Before = effect12$n1, 
                 Statistic = test12$statistic, 
                 #df = test12$parameter, 
                 EffectSize = effect12$effsize,
                 Pvalue = test12$p.value)
```

#### InvSimpson

```{r}
shapiro.test(df11$InvSimpson)
hist(df11$InvSimpson,probability=T, main="MS.TP1 vs MS.TP2",xlab="InvSimpson") 
lines(density(df11$InvSimpson),col=2)

```


```{r}
test13 <- wilcox.test(InvSimpson ~ Group, data = df11, paired = TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)

test13

```

```{r}
##effect size
effect13 <- wilcox_effsize(df11, InvSimpson ~ Group, paired = TRUE)
effect13

```

```{r}
#create tabl values

t13 <- data.frame(Group = "After(TP2) vs Before(TP1)", 
                 DataName = test13$data.name,
                 Method = test13$method,
                 After = effect13$n2, Before = effect13$n1, 
                 Statistic = test13$statistic, 
                 #df = test12$parameter, 
                 EffectSize = effect13$effsize,
                 Pvalue = test13$p.value)
```


## MS:TP1 vs MS.TP3
```{r}
#Prepare a data.frame for Treated group.

df2 = cbind(sample_data(psST3.rar)[, "Samples", drop=FALSE], richness32)
df2$Shannon <- richness31$Shannon
df2$InvSimpson <- richness33$InvSimpson
df2$Age <- sample_data(psST3.rar)$Age
df2$Sex <- sample_data(psST3.rar)$Sex
df2$Group <- sample_data(psST3.rar)$Group
df2$TimePoint <- sample_data(psST3.rar)$TimePoint
df2$ParticipantID <- sample_names(psST3.rar)
df2$logChao1 <- log(df2$Chao1)
df2$Treat <- sample_data(psST3.rar)$Ocrevus
df2 <- df2[c(1,3,4,5,6,7,2,8,9,10,11,12)]
df2

#samples are in each time point
table(df2$TimePoint)
```

```{r}
#select only duplicated elements
xu2 <- df2$Samples[duplicated(df2$Samples)]
xu2

df22 <- subset(df2, Samples %in% xu2)
df22

```


#### Chao1

```{r}
shapiro.test(df22$Chao1)
hist(df22$Chao1,probability=T, main="MS.TP1 vs MS.TP3",xlab="Chao1") 
lines(density(df22$Chao1),col=2)


```

```{r}
test2 <- t.test(Chao1 ~ Group, data = df22, paired = TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)

```

```{r}
##effect size
effect2 <- df22 %>% cohens_d(Chao1 ~ Group, var.equal = TRUE, hedges.correction = TRUE)

```

```{r}
#create tabl values

t2 <- data.frame(Group = "After(TP3) vs Before(TP1)", 
                 DataName = test2$data.name,
                 Method = test2$method,
                 After = effect2$n2, Before = effect2$n1, 
                 Statistic = test2$statistic, 
                 #df = test2$parameter, 
                 EffectSize = effect2$effsize,
                 Pvalue = test2$p.value)
```


#### Shannon

```{r}
shapiro.test(df22$Shannon)
hist(df22$Shannon,probability=T, main="MS.TP1 vs MS.TP3",xlab="Shannon") 
lines(density(df22$Shannon),col=2)

```


```{r}
test22 <- wilcox.test(Shannon ~ Group, data = df22, paired = TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)

test22

```

```{r}
##effect size

effect22 <- wilcox_effsize(df22, Shannon ~ Group, paired = TRUE)
effect22

```

```{r}
#create tabl values

t22 <- data.frame(Group = "After(TP3) vs Before(TP1)", 
                 DataName = test22$data.name,
                 Method = test22$method,
                 After = effect22$n2, Before = effect22$n1, 
                 Statistic = test22$statistic, 
                 #df = test12$parameter, 
                 EffectSize = effect22$effsize,
                 Pvalue = test22$p.value)
```

#### InvSimpson

```{r}
shapiro.test(df22$InvSimpson)
hist(df22$InvSimpson,probability=T, main="MS.TP1 vs MS.TP3",xlab="InvSimpson") 
lines(density(df22$InvSimpson),col=2)

```


```{r}
test23 <- wilcox.test(InvSimpson ~ Group, data = df22, paired = TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)

test23

```

```{r}
##effect size

effect23 <- wilcox_effsize(df22, InvSimpson ~ Group, paired = TRUE)

effect23

```

```{r}
#create tabl values

t23 <- data.frame(Group = "After(TP3) vs Before(TP1)", 
                 DataName = test23$data.name,
                 Method = test23$method,
                 After = effect23$n2, Before = effect23$n1, 
                 Statistic = test23$statistic, 
                 #df = test12$parameter, 
                 EffectSize = effect23$effsize,
                 Pvalue = test23$p.value)
```


## MS:TP1 vs MS.TP4
```{r}
#Prepare a data.frame for Treated group.

df3 = cbind(sample_data(psST4.rar)[, "Samples", drop=FALSE], richness42)
df3$Shannon <- richness41$Shannon
df3$InvSimpson <- richness43$InvSimpson
df3$Age <- sample_data(psST4.rar)$Age
df3$Sex <- sample_data(psST4.rar)$Sex
df3$Group <- sample_data(psST4.rar)$Group
df3$TimePoint <- sample_data(psST4.rar)$TimePoint
df3$ParticipantID <- sample_names(psST4.rar)
df3$logChao1 <- log(df3$Chao1)
df3$Treat <- sample_data(psST4.rar)$Ocrevus
df3 <- df3[c(1,3,4,5,6,7,2,8,9,10,11,12)]
df3

#samples are in each time point
table(df3$TimePoint)
```

```{r}
#select only duplicated elements
xu3 <- df3$Samples[duplicated(df3$Samples)]
xu3

df33 <- subset(df3, Samples %in% xu3)
df33

```

```{r}
#test for covariates

#Chi-square to check if factor are dependent
res <- compareGroups(Group ~ Age + Sex,data = df33, Q1=0.25, Q3=0.75 )

createTable(res)

```

####Chao1

```{r}
shapiro.test(df33$Chao1)
hist(df33$Chao1,probability=T, main="MS.TP1 vs MS.TP4",xlab="Chao1") 
lines(density(df33$Chao1),col=2)

```

```{r}
test3 <- t.test(Chao1 ~ Group, data = df33, paired = TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)

```

```{r}
##effect size
effect3 <- df33 %>% cohens_d(Chao1 ~ Group, var.equal = TRUE, hedges.correction = TRUE)

```

```{r}
#create tabl values

t3 <- data.frame(Group = "After(TP4) vs Before(TP1)", 
                 DataName = test3$data.name,
                 Method = test3$method,
                 After = effect3$n2, Before = effect3$n1, 
                 Statistic = test3$statistic, 
                 #df = test3$parameter, 
                 EffectSize = effect3$effsize,
                 Pvalue = test3$p.value)
```

#### Shannon

```{r}
shapiro.test(df33$Shannon)
hist(df33$Shannon,probability=T, main="MS.TP1 vs MS.TP4",xlab="Shannon") 
lines(density(df33$Shannon),col=2)

```


```{r}
test32 <- wilcox.test(Shannon ~ Group, data = df33, paired = TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)

test32

```

```{r}
##effect size

effect32 <- wilcox_effsize(df33, Shannon ~ Group, paired = TRUE)
effect32

```

```{r}
#create tabl values

t32 <- data.frame(Group = "After(TP4) vs Before(TP1)", 
                 DataName = test32$data.name,
                 Method = test32$method,
                 After = effect32$n2, Before = effect32$n1, 
                 Statistic = test32$statistic, 
                 #df = test12$parameter, 
                 EffectSize = effect32$effsize,
                 Pvalue = test32$p.value)
```

#### InvSimpson

```{r}
shapiro.test(df33$InvSimpson)
hist(df33$InvSimpson,probability=T, main="MS.TP1 vs MS.TP4",xlab="InvSimpson") 
lines(density(df33$InvSimpson),col=2)

```


```{r}
test33 <- wilcox.test(InvSimpson ~ Group, data = df33, paired = TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)

test33

```

```{r}
##effect size

effect33 <- wilcox_effsize(df33, InvSimpson ~ Group, paired = TRUE)

effect33

```

```{r}
#create tabl values

t33 <- data.frame(Group = "After(TP4) vs Before(TP1)", 
                 DataName = test33$data.name,
                 Method = test33$method,
                 After = effect33$n2, Before = effect33$n1, 
                 Statistic = test33$statistic, 
                 #df = test12$parameter, 
                 EffectSize = effect33$effsize,
                 Pvalue = test33$p.value)
```


# Final models

```{r}
#merge tables
before.after <- rbind(t1, t12, t13, t2, t22, t23, t3, t32, t33)
```

# save output

```{r}
#export(before.after, "MS.stool.before.after.diversity.oc.xlsx")
```


# Session info

