---
title: "Diversity Analysis: MS-O vs HC.stool; MS-nO vs HC.stool"
author: "Alba Troci"
date: "`r Sys.Date()`"  
output:
   BiocStyle::html_document:
      toc: true
      df_print: paged
      self_contained: true
      code_download: true
      highlight: tango
editor_options:
  chunk_output_type: console
params:
  FIGPATH: "figures/"
  d.out: "./"
---

# Preparations
## Set global options

```{r style, results="asis", cache=FALSE, message = F, echo=FALSE}
# Set knit global options
library("knitr")
options(digits = 2, width = 80)
golden_ratio <- (1 + sqrt(5)) / 2
opts_chunk$set(echo = TRUE,
               tidy = FALSE,
               include = TRUE,
               fig.path = params$FIGPATH,
               dev=c("png",'pdf'),
               fig.height = 5,
               fig.width = 4 * golden_ratio,
               comment = '  ',
               dpi = 300,
               cache = FALSE)


# Pretty outputs
library("rmarkdown")
library("BiocStyle")
library("ggthemes") # add ggplot2 themes
library("ggplot2")
library("ggpubr")

# Set seed for reproducibility
set.seed(100)

# Set plotting theme
theme_set(theme_few(base_size = 14))

# Set output directory
d.out <- params$d.out
rm(params)
```


## Load libraries for the session
```{r libraries, message=FALSE}
library("magrittr")
library("ggplot2")
library("tidyverse")
library("readxl")
library("dada2")
library("phyloseq")
library("gridExtra")
library("vegan")
library("plyr")
library("scales")
library("reshape2")
library("DESeq2")
library("dplyr") 
library("msa") 
library("ape") 
library("randomForest") 
library("caret") 
library("broom")
library("mlbench")
library("plotROC")
library("car")
library("MASS")
library("lme4")
library("lmerTest")
library("fitdistrplus")
library("psycho")
library("rstatix")
library("coin")
library("plotROC")
library("gg3D")
library("scatterplot3d")
library("car")
library("rgl")
library("rglwidget")
library("rstatix")
library("coin")
library("compareGroups")
```

## Import data: no agglomeration

```{r import data}
# Phyloseq object: Filtered ASV based from MS+Healthy samples 
d.ps1_HMS <-"~/IKMB/projects/MS/github.codes.ms/ms.vs.hc/data"
ps1_HMS <- "ps_HMS.subset.stool.itm.rds" %>%
  paste0(d.ps1_HMS,"/", .) %>% 
  readRDS() 
ps1_HMS

```


## Edit
```{r}
dim(sample_data(ps1_HMS)[sample_data(ps1_HMS)$Status %in% "Healthy"])
dim(sample_data(ps1_HMS)[sample_data(ps1_HMS)$Status %in% "MS"])
dim(sample_data(ps1_HMS)[sample_data(ps1_HMS)$Ocrevus %in% "yes"])
dim(sample_data(ps1_HMS)[sample_data(ps1_HMS)$Ocrevus %in% "no"])
```

```{r}
sample_data(ps1_HMS)$TimePoint[sample_data(ps1_HMS)$Status %in% "Healthy"] <- "TP0"

sample_data(ps1_HMS)$Status[sample_data(ps1_HMS)$Status %in% "Healthy"] <- "HC"
sample_data(ps1_HMS)$Status[sample_data(ps1_HMS)$Ocrevus %in% "yes"] <- "MS-O"
sample_data(ps1_HMS)$Status[sample_data(ps1_HMS)$Ocrevus %in% "no"] <- "MS-nO"

```

```{r}
ps.hms <- subset_samples(ps1_HMS, Status %in% c("HC", "MS-O", "MS-nO"))
ps.hms

sam <- data.frame(sample_data(ps.hms))

```


# Alpha diversity test

## Rarefaction
We need to rarefy before alpha diversity:

```{r, message=FALSE}
# We will subsample to the min number of reads
summary(sample_sums(ps.hms))
sample_sums(ps.hms)
```


```{r}
ps.hms
ps_merge.rar <- rarefy_even_depth(ps.hms, sample.size = min(sample_sums(ps.hms)))
ps_merge.rar

```

Subset only HC and MS in baseline

```{r}
ps.h <- subset_samples(ps_merge.rar, Status == "HC")
ps.h

ps.1 <- subset_samples(ps_merge.rar, TimePoint %in% c("TP0", "TP1"))
ps.1
```


```{r}
theme_set(theme_bw())

#estimate shannon alpha diversity measure
alpha1 <- estimate_richness(ps.1, split = TRUE, measures = "shannon")

df = cbind(sample_data(ps.1)[, "Age", drop=FALSE], alpha1)
df$Sex <- sample_data(ps.1)$Sex
df$Status <- sample_data(ps.1)$Status
df
```

Create separate tables for MS-O and MS-nO
```{r}
df1 <- df[(df$Status %in% c("HC", "MS-O")), ]
df2 <- df[(df$Status %in% c("HC", "MS-nO")), ]
```

Test for Covariates 
```{r}
#For MS-O vs HC
#Chi-square to check if factor are dependent
res <- compareGroups(Status ~ Age + Sex,data = df1, Q1=0.25, Q3=0.75 )
createTable(res)

#For MS-nO vs HC
#Chi-square to check if factor are dependent
res2 <- compareGroups(Status ~ Age + Sex,data = df2, Q1=0.25, Q3=0.75 )
createTable(res2)
```

Test if Shannon values are normally distributed
```{r}
#For MS-O vs HC
hist(df1$Shannon,probability=T, main="MS-O vs Healthy",xlab="Shannon") 
lines(density(df1$Shannon),col=2)
shapiro.test(df1$Shannon)


#For MS-O vs HC
hist(df2$Shannon,probability=T, main="MS-nO vs Healthy",xlab="Shannon") 
lines(density(df2$Shannon),col=2)
shapiro.test(df2$Shannon)
```

Test relationship between Shannon alpha diversity measure and Status (MS-O vs HC)
```{r}
#shannon
w1 <- wilcox.test(Shannon ~ Status, data = df1,conf.int=TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)
w1

#df1 %>% cohens_d(Shannon ~ Status, var.equal = TRUE, hedges.correction = TRUE)
effect1 <- df1 %>% wilcox_effsize(Shannon ~ Status)
effect1
```

```{r}
#create tabl values

t.Shannon1 <- data.frame(Status = "MS-O vs HC", 
                 DataName = w1$data.name,
                 Method = w1$method,
                 MS.Baseline = effect1$n2, HC = effect1$n1, 
                 Statistic = w1$statistic, 
                 EffectSize = effect1$effsize,
                 Pvalue = w1$p.value)
```


Test relationship between Shannon alpha diversity measure and Status (MS-nO vs HC)
```{r}
# wilcox. test for Shannon: evenness
compare2 <- pairwise.wilcox.test(df2$Shannon, 
                                 df2$Status,
                                 alternative = "two.sided",
                                 p.adjust.method = "bonferroni",exact=FALSE)
compare2
```

```{r}
#shannon
w2 <- wilcox.test(Shannon ~ Status, data = df2,conf.int=TRUE,
       p.adjust.methods = "bonferroni",var.equal = TRUE)
w2

#df1 %>% cohens_d(Shannon ~ Status, var.equal = TRUE, hedges.correction = TRUE)
effect2 <- df2 %>% wilcox_effsize(Shannon ~ Status)
effect2
```

```{r}
#create tabl values

t.Shannon2 <- data.frame(Status = "MS-nO vs HC", 
                 DataName = w2$data.name,
                 Method = w2$method,
                 MS.Baseline = effect2$n2, HC = effect2$n1, 
                 Statistic = w2$statistic, 
                 EffectSize = effect2$effsize,
                 Pvalue = w2$p.value)
```


# Alpha diversity plot

Prepare data table for ggplot
```{r}
df200 <- df[c("Shannon", "Status")]
df200$Value <- df200$Shannon
df200$Measure <- "Shannon"
df200$Shannon <- NULL

df200$Group[df200$Status %in% "HC"] <- "HC"
df200$Group[df200$Status %in% "MS-O"] <- "MS"
df200$Group[df200$Status %in% "MS-nO"] <- "MS"

colnames(df200) <- c("Group" ,  "Value" ,  "Measure", "Status" )
```

Add specific colors
```{r}
cols <- c("royalblue4", "tomato4", "orange3")
```

Plot
```{r}
theme_set(theme_bw())

p <- ggplot(df200, aes(x=Group, y=Value, color=Group)) +
  geom_boxplot(alpha = 2, na.rm = TRUE, outlier.size = 0)+ 
                        
                       theme(text = element_text(size = 15)) +
                       scale_colour_manual(values = cols) +
                       labs(y="Alpha diversity measure", x = "Group") +
                       scale_y_continuous(expand = expansion(mult = c(0, 0.4)))

hms.baseline.stool.shannon <- p + facet_wrap(~Measure, scale="free")
hms.baseline.stool.shannon 
#ggsave("hms.baseline.stool.shannon.space.png")
```



# Session info


